---
title: "Heart Disease Prediction"
author: "Aadesh Anil Nikam"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 9,
  fig.height = 6
)
```




```{r packages}
pkgs <- c(
  "tidyverse", "tableone", "corrplot", "caret",
  "pROC", "glmnet", "randomForest",
  "ResourceSelection", "dcurves", "gridExtra", "broom"
)
new_pkgs <- pkgs[!pkgs %in% installed.packages()[, "Package"]]
if (length(new_pkgs)) install.packages(new_pkgs, dependencies = TRUE)
invisible(lapply(pkgs, library, character.only = TRUE))

set.seed(42)
```



## Load Data

```{r load-data}
heart_raw <- read.csv("heart.csv", stringsAsFactors = FALSE)
cat("Raw dimensions:", nrow(heart_raw), "rows x", ncol(heart_raw), "cols\n")
```

## Duplicate Rows

```{r}
n_dup <- sum(duplicated(heart_raw))
cat("Exact duplicate rows found:", n_dup, "\n")
cat("These are removed to avoid artificially inflated test-set performance.\n")

heart <- distinct(heart_raw)
cat("Rows after deduplication:", nrow(heart), "\n")
```

## Missing Values

```{r}
cat("Missing values per variable:\n")
print(colSums(is.na(heart)))
cat("Total missing:", sum(is.na(heart)), "\n")
```

## Recode Variables

```{r}
heart <- heart %>%
  mutate(
    target  = factor(target,  levels = c(0, 1),
                     labels = c("No_HD", "HD")),
    sex     = factor(sex,     levels = c(0, 1),
                     labels = c("Female", "Male")),
    cp      = factor(cp,      levels = 0:3,
                     labels = c("Typical", "Atypical", "NonAnginal", "Asympt")),
    fbs     = factor(fbs,     levels = c(0, 1),
                     labels = c("Normal", "High")),
    restecg = factor(restecg, levels = 0:2,
                     labels = c("Normal", "ST_T", "LV_Hypertrophy")),
    exang   = factor(exang,   levels = c(0, 1),
                     labels = c("No", "Yes")),
    slope   = factor(slope,   levels = 0:2,
                     labels = c("Down", "Flat", "Up")),
    ca      = factor(ca),
    thal    = factor(thal,    levels = 0:3,
                     labels = c("Unknown", "Normal", "Fixed", "Reversible"))
  )
```

## Outcome Distribution

```{r}
cat("Outcome distribution (deduplicated sample):\n")
print(table(heart$target))
print(round(prop.table(table(heart$target)), 3))
```

## Data Size (n × p)

```{r size}
cat("Observations (n):", nrow(heart), "\n")
cat("Predictors (p):  ", ncol(heart) - 1, "\n")
cat("HD cases:        ", sum(heart$target == "HD"), "\n")
cat("Non-HD controls: ", sum(heart$target == "No_HD"), "\n")
```


# Exploratory Data Analysis

```{r define-vars}
cat_vars  <- c("sex", "cp", "fbs", "restecg", "exang", "slope", "ca", "thal")
cont_vars <- c("age", "trestbps", "chol", "thalach", "oldpeak")
```

## Table 1 — Descriptive Statistics by Outcome

```{r}
tab1 <- CreateTableOne(
  vars       = c(cont_vars, cat_vars),
  strata     = "target",
  data       = heart,
  factorVars = cat_vars
)
print(tab1, showAllLevels = TRUE, quote = FALSE, noSpaces = TRUE)
```

## Continuous Predictors by Outcome

```{r}
cont_long <- heart %>%
  select(target, all_of(cont_vars)) %>%
  pivot_longer(-target, names_to = "variable", values_to = "value")

ggplot(cont_long, aes(x = value, fill = target)) +
  geom_histogram(alpha = 0.5, bins = 30, position = "identity") +
  facet_wrap(~variable, scales = "free") +
  scale_fill_manual(values = c("steelblue", "firebrick"),
                    name = "Heart Disease") +
  labs(title = "Continuous Predictors by Heart Disease Status",
       x = NULL, y = "Count") +
  theme_bw(base_size = 11)
```

## Categorical Predictors by Outcome

```{r}
cat_long <- heart %>%
  select(target, all_of(cat_vars)) %>%
  pivot_longer(-target, names_to = "variable", values_to = "level")

ggplot(cat_long, aes(x = level, fill = target)) +
  geom_bar(position = "fill") +
  facet_wrap(~variable, scales = "free_x") +
  scale_fill_manual(values = c("steelblue", "firebrick"),
                    name = "Heart Disease") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Proportion with Heart Disease by Categorical Predictor",
       x = NULL, y = "Proportion") +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

## Correlation Matrix (Continuous Variables Only)

```{r}
corrplot(
  cor(heart %>% select(all_of(cont_vars))),
  method = "color", type = "upper",
  addCoef.col = "black", tl.cex = 0.9, number.cex = 0.85,
  title = "Correlation Matrix — Continuous Predictors",
  mar   = c(0, 0, 2, 0)
)
```


# Train / Test Split

```{r}
train_idx <- createDataPartition(heart$target, p = 0.70, list = FALSE)
train_df  <- heart[ train_idx, ]
test_df   <- heart[-train_idx, ]

cat("Training set:", nrow(train_df), "rows |",
    sum(train_df$target == "HD"), "HD cases\n")
cat("Test set:    ", nrow(test_df),  "rows |",
    sum(test_df$target  == "HD"), "HD cases\n")

# Numeric outcome for test set (used in calibration / DCA)
y_test <- as.numeric(test_df$target == "HD")
```


# Methods & Results

## Helper Functions

```{r}
# Predicted class from probabilities
make_class <- function(probs, thr = 0.5) {
  factor(ifelse(probs > thr, "HD", "No_HD"), levels = c("No_HD", "HD"))
}


# Performance metrics
get_metrics <- function(probs, actual, model_name) {
  cm      <- confusionMatrix(make_class(probs), actual, positive = "HD")
  roc_obj <- roc(actual, probs, levels = c("No_HD", "HD"), quiet = TRUE)
  brier   <- mean((probs - as.numeric(actual == "HD"))^2)
  data.frame(
    Model       = model_name,
    AUC         = round(auc(roc_obj), 4),
    Accuracy    = round(cm$overall["Accuracy"],       4),
    Sensitivity = round(cm$byClass["Sensitivity"],    4),
    Specificity = round(cm$byClass["Specificity"],    4),
    PPV         = round(cm$byClass["Pos Pred Value"], 4),
    NPV         = round(cm$byClass["Neg Pred Value"], 4),
    Brier       = round(brier, 4),
    row.names   = NULL
  )
}
```



## Model 1 — Logistic Regression

Logistic regression was fitted using all 13 predictors. It serves as the interpretable baseline model, providing odds ratios that have direct clinical meaning.

```{r}
lr_fit <- glm(target ~ ., data = train_df, family = binomial)
summary(lr_fit)
```

### Odds Ratios

```{r}
round(exp(cbind(OR = coef(lr_fit), confint(lr_fit))), 3)
```

### Coefficient Plot

```{r}
coef_df <- tidy(lr_fit, conf.int = TRUE, exponentiate = FALSE) %>%
  filter(term != "(Intercept)") %>%
  mutate(direction = ifelse(estimate > 0, "Increases risk", "Decreases risk"))

ggplot(coef_df,
       aes(x = reorder(term, estimate), y = estimate,
           colour = direction, ymin = conf.low, ymax = conf.high)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(size = 0.7) +
  coord_flip() +
  scale_colour_manual(values = c("firebrick", "steelblue"), name = NULL) +
  labs(title = "Logistic Regression — Log Odds Ratios with 95% CI",
       x = NULL, y = "Log Odds Ratio") +
  theme_bw(base_size = 11)
```

```{r}
pred_lr <- predict(lr_fit, newdata = test_df, type = "response")
roc_lr  <- roc(test_df$target, pred_lr,
               levels = c("No_HD", "HD"), quiet = TRUE)
cat("Logistic Regression AUC:", round(auc(roc_lr), 4), "\n")
```


## Model 2 — LASSO Logistic Regression

LASSO applies an L1 penalty to shrink unimportant coefficients to zero, performing automatic variable selection. The penalty parameter lambda was chosen by 10-fold cross-validation maximising AUC.

```{r}
dummies <- dummyVars(target ~ ., data = train_df, fullRank = TRUE)
X_train <- predict(dummies, newdata = train_df)
X_test  <- predict(dummies, newdata = test_df)

cv_lasso <- cv.glmnet(
  X_train,
  as.numeric(train_df$target == "HD"),
  family       = "binomial",
  alpha        = 1,
  nfolds       = 10,
  type.measure = "auc"
)
```

### Cross-Validation Curve

```{r}
plot(cv_lasso, main = "LASSO — Cross-Validated AUC vs log(lambda)")
cat("Best lambda (1-SE rule):", round(cv_lasso$lambda.1se, 5), "\n")
```

### Selected Predictors

```{r}
lasso_fit <- glmnet(
  X_train,
  as.numeric(train_df$target == "HD"),
  family = "binomial",
  alpha  = 1,
  lambda = cv_lasso$lambda.1se
)

lasso_coef <- coef(lasso_fit)
cat("Non-zero (selected) coefficients:\n")
print(lasso_coef[lasso_coef[, 1] != 0, , drop = FALSE])
```

```{r lasso-pred}
pred_lasso <- as.vector(predict(lasso_fit, newx = X_test, type = "response"))
roc_lasso  <- roc(test_df$target, pred_lasso,
                  levels = c("No_HD", "HD"), quiet = TRUE)
cat("LASSO AUC:", round(auc(roc_lasso), 4), "\n")
```



## Model 3 — Random Forest

Random Forest grows 500 decision trees on bootstrap samples, each using a random subset of √p predictors at each split. It captures non-linear relationships and interactions without requiring variable transformations.

```{r}
rf_fit <- randomForest(
  target ~ .,
  data       = train_df,
  ntree      = 500,
  mtry       = floor(sqrt(ncol(train_df) - 1)),
  importance = TRUE
)
print(rf_fit)
```

### Variable Importance

```{r}
varImpPlot(rf_fit, main = "Random Forest — Variable Importance")
```

```{r}
pred_rf <- predict(rf_fit, newdata = test_df, type = "prob")[, "HD"]
roc_rf  <- roc(test_df$target, pred_rf,
               levels = c("No_HD", "HD"), quiet = TRUE)
cat("Random Forest AUC:", round(auc(roc_rf), 4), "\n")
```


# Model Evaluation

## ROC Curves & AUC

> AUC and ROC curves are **fully valid** in a case-control dataset because they measure discrimination — the ability to rank cases above controls — and are not affected by the case-control sampling fraction.

### Individual ROC Curves

```{r}
par(mfrow = c(1, 3))

plot(roc_lr,    col = "royalblue",   lwd = 2,
     main = paste0("Logistic Regression\nAUC = ", round(auc(roc_lr), 3)))
abline(a = 1, b = -1, lty = 2, col = "grey60")

plot(roc_lasso, col = "darkorange",  lwd = 2,
     main = paste0("LASSO\nAUC = ", round(auc(roc_lasso), 3)))
abline(a = 1, b = -1, lty = 2, col = "grey60")

plot(roc_rf,    col = "forestgreen", lwd = 2,
     main = paste0("Random Forest\nAUC = ", round(auc(roc_rf), 3)))
abline(a = 1, b = -1, lty = 2, col = "grey60")

par(mfrow = c(1, 1))
```

### Combined ROC Overlay

```{r}
plot(roc_lr,    col = "royalblue",   lwd = 2,
     main = "ROC Curves — All Models (Test Set)")
plot(roc_lasso, col = "darkorange",  lwd = 2, add = TRUE)
plot(roc_rf,    col = "forestgreen", lwd = 2, add = TRUE)
abline(a = 1, b = -1, lty = 2, col = "grey60")

legend("bottomright", bty = "n",
  legend = c(
    paste0("Logistic Regression  AUC = ", round(auc(roc_lr),    3)),
    paste0("LASSO                AUC = ", round(auc(roc_lasso),  3)),
    paste0("Random Forest        AUC = ", round(auc(roc_rf),     3))
  ),
  col = c("royalblue", "darkorange", "forestgreen"),
  lwd = 2
)
```

### AUC Comparison Tests

```{r}
dl_lr_lasso <- roc.test(roc_lr,    roc_lasso, method = "delong")
dl_lr_rf    <- roc.test(roc_lr,    roc_rf,    method = "delong")
dl_lasso_rf <- roc.test(roc_lasso, roc_rf,    method = "delong")

cat(sprintf("LR vs LASSO:         p = %.4f\n", dl_lr_lasso$p.value))
cat(sprintf("LR vs Random Forest: p = %.4f\n", dl_lr_rf$p.value))
cat(sprintf("LASSO vs RF:         p = %.4f\n", dl_lasso_rf$p.value))
```
## Confusion Matrices

> A classification threshold of 0.5 is used here for illustration. Note that this threshold may not be optimal clinically — in a real application, the threshold would be chosen based on the relative costs of false positives and false negatives.

```{r}
cat("--- Logistic Regression ---\n")
print(confusionMatrix(make_class(pred_lr),    test_df$target, positive = "HD"))

cat("--- LASSO ---\n")
print(confusionMatrix(make_class(pred_lasso), test_df$target, positive = "HD"))

cat("--- Random Forest ---\n")
print(confusionMatrix(make_class(pred_rf),    test_df$target, positive = "HD"))
```


## Decision Curve Analysis

```{r}
dca_df <- data.frame(
  target = y_test,
  lr     = pred_lr,
  lasso  = pred_lasso,
  rf     = pred_rf
)

dca_res <- dca(
  target ~ lr + lasso + rf,
  data       = dca_df,
  thresholds = seq(0.05, 0.95, by = 0.01),
  label      = list(
    lr    = "Logistic Regression",
    lasso = "LASSO",
    rf    = "Random Forest"
  )
)

plot(dca_res, smooth = TRUE) +
  labs(
    title    = "Decision Curve Analysis — Test Set (illustrative only)",
    subtitle = "Interpret cautiously: case-control data does not reflect true population prevalence"
  ) +
  theme_bw(base_size = 12)
```
## Performance Summary Table

```{r}
perf_table <- bind_rows(
  get_metrics(pred_lr,    test_df$target, "Logistic Regression"),
  get_metrics(pred_lasso, test_df$target, "LASSO"),
  get_metrics(pred_rf,    test_df$target, "Random Forest")
) %>% arrange(desc(AUC))

knitr::kable(perf_table,
             caption = "Model Performance on Test Set (AUC, discrimination metrics fully valid; Brier score affected by case-control design)",
             digits  = 4,
             align   = "lrrrrrrr")
```

# Prevalence Adjustment for Predicted Probabilities

Because this is a case-control dataset, raw predicted probabilities are calibrated to the **sample event rate (\~50%)**, not the true population prevalence. To obtain meaningful absolute risk estimates for clinical use
> We demonstrate this recalibration using an **assumed population prevalence of 7%** for illustration. In a real application, this figure should be obtained from a reliable external source (e.g., national registry data).

```{r}
pi_sample <- mean(y_test)   
pi_pop    <- 0.07           

offset <- qlogis(pi_pop) - qlogis(pi_sample)

pred_lr_recal <- plogis(qlogis(pred_lr) + offset)

cat("Sample event rate (case-control):  ", round(pi_sample, 3), "\n")
cat("Assumed population prevalence:     ", pi_pop, "\n")
cat("Log-odds offset applied:           ", round(offset, 4), "\n\n")
cat("Mean predicted prob — raw:         ", round(mean(pred_lr), 3), "\n")
cat("Mean predicted prob — recalibrated:", round(mean(pred_lr_recal), 3), "\n")
```






